To be FAIR: Theory Specification Needs an Update

### 2025-06-19 test.Rmd
update 2025-06-19 15:08:14.390215

authornote: |
  We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion. We also thank the Editor, Dr. Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments. Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #`r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

<span id="whatsfair">The FAIR Guiding Principles (hereafter: FAIR principles) were established by a diverse consortium of stakeholders to improve the reusability of research data and other resources produced during the course of scholarly work
by making them Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016]^[[As the colloquial use of these terms differs from their definition according to the FAIR principles, we capitalize these terms when referring to specific FAIR principles.]{#faircap}].
Since the FAIR principles' inception, 
they have become a widely adopted standard for archival of academic output,
representing an estimated tens of billions of dollars in reuse value [@vogtFAIR20Extending2024].
Scholars have demonstrated their relevance for making other digital objects more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].</span>
The present paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of scientific theory, published as a self-contained and citable digital object distinct from - but potentially connected to - the scientific paper.
<span id="alldefinitions">Definitions of theory are abundant and hotly debated,
but many of them are compatible with the FAIR principles.</span>
FAIR theory can potentially improve the transparency and efficiency of scholarly communication, reduce research waste, and accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
rendering them immune to falsification.
As an example, consider "self-determination theory" (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Intrinsic motivation was initially defined as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, this definition expanded to include motivations driven by the fulfillment of basic psychological needs for autonomy, competence, and relatedness [@ryanSelfdeterminationTheoryFacilitation2000].
<!-- Consider, for instance, a research scientist working in a highly competitive laboratory.
Under the original definition, their intrinsic motivation would be limited to the joy of discovery alone.
Yet, under the expanded framework, factors that might initially appear extrinsic can be reconceptualized as intrinsic: securing grant funding enhances their autonomy by granting them the freedom to pursue innovative ideas; grappling with deadlines bolsters their sense of competence; and praise from their peers fulfills their need for relatedness. -->
The implications of these shifting definitions becomes clear when deriving hypotheses about the type of motivation involved in changing an infant's dirty diaper.
Under the original definition, one would hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one would hypothesize the opposite, as the act fulfills the need for relatedness.
The expanded definition thus enables SDT to absorb potentially falsifying evidence.

Scholars have raised concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One frequently raised concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
When theories are ambiguous,
precise predictions cannot be derived from them without
resorting to subjective interpretation or invoking additional assumptions,
which makes them harder to falsify.
A second concern has received less attention,
is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical processes of theory construction and improvement have been largely overlooked.
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We apply introduce *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

Given that a pluriformity of definitions are consistent with FAIR theory principles,
we do not limited ourselves to any one particular definition - although, at times, our writing will inevitably reveal a particular vantage point.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain (Table \@ref(tab:tabmeehl)):
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [@frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
models have been described as "specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020, p. 336].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.

@@

Perhaps some ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are not of substantive interest, but merely reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
such as straw-man null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeachersCornerEvaluating2020].
<span id = "waste">Testing ad-hoc hypotheses not grounded in theory, or grounded in misinterpreted- or multi-interpretable theory,
does not advance our principled understanding of psychological phenomena,
and consequently contributes to research waste [@nakagawaPoorHypothesesResearch2024].</span>
<span id = "norms">Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021]; inefficient at best, and more likely, futile.
As the Declaration of Helsinki prescribes that ethical (medical) research with human participants must "avoid research waste",
our field should take seriously its ethical responsibility to develop procedures to reduce it.</span>
The present paper does so by introducing procedures to improve transparent and unambiguous communication about theory; 
instantiating theory as a digital "object" that scholars can access, reuse, and update in their daily workflows.

@@

Merely publishing theory in a journal article does not make it open;
to be open, theory should adhere to established open science standards for specification and archival.
We propose to implement theories as digital objects,
and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data". 
They provide information about the nature and content of a digital object and are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposes, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues [-@lamprechtFAIRPrinciplesResearch2019s],
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary (minor) changes,
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR digital object,
and we provide worked examples to encourage their adoption.

@@

It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be archived as FAIR theory. 
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind of information is useful.
As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
For example, following Meehl, we could envision a theory that starts out with establishing, through observation, an ontology of constructs relevant for a given phenomenon.
After initial exploratory research, the theory might be further specified by making assumptions about how these constructs are causally connected.
Over time, more precise *statistical/mathematical models* could be derived by further assuming a specific functional form for relationships (e.g., linear effects) and error families for the distribution of measured variables (e.g., normal distributions).
This allows for the specification of statistical models, which make just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Going even further, a *generative/computational model* could be specified,
which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement tools (e.g., different questionnaires used to assess the same construct), or study subjects (e.g., specific strains of rats).

As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
such as adherence to infection prevention protocols (e.g., social distancing),
pandemic-related behavior (e.g., panic buying),
and pandemic-related distress
[@taylor2022psychology].
The theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.

@@

Crucially for the present paper, the issue of formalization is orthogonal to FAIRification.
FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it pertains to rigorous and transparent archival and communication about theories, with the aim of enhancing their reusability.
FAIR theory does not require formalization, and formal theories are not automatically FAIR.
The FAIR principles
apply to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).

@@

We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
<span id = "modularvalue">The primary unit of scientific communication has long been the academic paper.
In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
These resources are often merely described in papers (if at all), and not made available for reuse.
Modular publishing is the practice of making each of these resources available as independent *digital objects*,
facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].</span>
<span id="datasharing">Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology [@tedersooDataSharingPractices2021].</span>

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving digital objects.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
Note that archival of the version of record is *not* a function of Git(Hub).
While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.

@@

Aside from technical solutions, version control is a social process as well.
On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
<!-- NvD: Maybe we can add an  example (e.g., the theory that explains ego-depletion and the failed replication of the phenomenon). -->
In computer science, these challenges are navigated by assigning version numbers.
Specifically, *semantic versioning* comprises a simple set of rules for assigning version numbers to digital objects.
Whereas version control tracks changes,
semantic versioning communicates what those changes mean to users of the theory,
guides the social process of theory development, and signals how much a theory has been changed.
We propose the following adaptation of semantic versioning for theories:

Given a version number in the format MAJOR.MINOR.PATCH (where MAJOR, MINOR, and PATCH are placeholders for positive integer numbers including zero), increment the:

- MAJOR version when you commit backwards incompatible changes, i.e., the
      theory now contains empirical statements that are at odds with
      a previous version of the theory
- MINOR version when you expand the set of empirical statements in a
      backward compatible manner (i.e., the previous version is
      subsumed within the new version)
- PATCH version when you make backward compatible bug fixes, cosmetic
      changes, fix spelling errors, or add clarifications

@@

The four Findability criteria are applicable to theory with only minor adjustments, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for archived objects,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata	(F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
ontology [@guyonMeasurementOntologyEpistemology2018],
or catalog of psychological phenomena.
Metadata should also include identifiers for all the versions of the theory it describes	(F3);
Zenodo handles this by default by providing an overarching DOI for a digital object which subsumes the DOIs of that artifact's versions.

Finally, metadata should be registered or indexed in a searchable registry (F4).
It is important to note that, while many archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
only few are specifically designed for FAIR-compliant archival.
Zenodo stands out in this respect.
Thus, while using Git for version control and GitHub for collaboration has specific advantages for scientific work [@ramGitCanFacilitate2013],
the version of record should be archived in a FAIR repository like Zenodo.
Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a specific) FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

Findability is substantially amplified if intended users of a resource know where to search for it.
This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
Regrettably, most academic search engines only index traditional print publications, not other digital objects.
Since the status quo is to publish theories in papers,
the FAIR requirements are met if scholars continue to do so,
and additionally publish theories as separate digital objects.
The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
<!-- An ad-hoc solution is to publish a paper as documentation for the digital object [@mcgillivrayDeepImpactStudy2022]. -->
<!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
<!-- The same solution could be applied to theories. -->
In the longer term, it may not be necessary to write a paper for each theory.
If Zenodo becomes more recognized as a centralized repository for digital objects, researchers may begin to search there more regularly.
Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index digital objects stored in Zenodo.

>>>>

authornote: |
  We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion. We also thank the Editor, Dr. Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments. Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #`r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized.

@@

<span id="whatsfair">The FAIR Guiding Principles (hereafter: FAIR principles) were established by a diverse consortium of stakeholders to improve the reusability of research data and other resources produced in the course of scholarly work
by making them Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016]^[[As the colloquial use of these terms differs from their definition according to the FAIR principles, we capitalize these terms when referring to specific FAIR principles.]{#faircap}].
Since the FAIR principles' inception, 
they have become a widely adopted standard for archival of academic output,
representing an estimated tens of billions of dollars in reuse value [@vogtFAIR20Extending2024].
Scholars have demonstrated their relevance for making other digital objects more open, including research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying; @vanlissaWORCSWorkflowOpen2021].</span>
The present paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of scientific theory, published as a self-contained and citable digital object distinct from - but potentially associated with - the scientific paper.
<span id="alldefinitions">Definitions of theory abound and hotly debated,
but as many of them are compatible with the FAIR principles,
this paper is not limited to a specific definition.</span>
FAIR theory can potentially improve the transparency and efficiency of scholarly communication, reduce research waste, and accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
rendering them immune to falsification.
Consider "self-determination theory" [SDT, @deciSelfDeterminationTheory2012], one of the most widely cited social psychological theories, which we formalized and FAIRified in [this vignette](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html).
SDT emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Intrinsic motivation was initially defined as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, this definition expanded to include motivations driven by the fulfillment of basic psychological needs for autonomy, competence, and relatedness [@ryanSelfdeterminationTheoryFacilitation2000].
The implications of these shifting definitions becomes clear when deriving hypotheses about the type of motivation involved in changing an infant's dirty diaper.
Under the original definition, one would hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one would hypothesize the opposite, as the act fulfills the need for relatedness.
Expanding the definition thus enables SDT to absorb potentially falsifying evidence.

Scholars have raised concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One frequently raised concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
When theories are ambiguous,
precise predictions cannot be derived from them without
resorting to subjective interpretation or invoking additional assumptions,
which makes them harder to falsify.
A second concern has received less attention,
is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical processes of theory construction and improvement have been largely overlooked.
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We introduce *FAIR theory* as a methodology that can facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

Given that a pluriformity of definitions are consistent with FAIR theory principles,
we do not limited ourselves to any one particular definition - although, at times, our writing inevitably reveals a particular vantage point.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain (Table \@ref(tab:tabmeehl)):
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [@frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
models have been described as "specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020, p. 336].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.

@@

Perhaps some ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be useful to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are not of substantive interest, but merely reported as part of entrenched cultural practices [@gigerenzerNullRitualWhat2004],
such as straw-man null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeachersCornerEvaluating2020].
<span id = "waste">Testing ad-hoc hypotheses not grounded in theory, or grounded in misinterpreted- or multi-interpretable theory,
cannot advance our principled understanding of psychological phenomena,
and consequently contributes to research waste [@nakagawaPoorHypothesesResearch2024].
<span id = "norms">Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021]; inefficient at best, and more likely, futile.
As the Declaration of Helsinki prescribes that ethical (medical) research with human participants must "avoid research waste",
our field should take seriously its ethical responsibility to develop procedures to reduce it.</span></span>
The present paper does so by introducing procedures to improve transparent and unambiguous communication about theory; 
instantiating theory as a digital "object" that scholars can access, reuse, and update in their daily workflows.

@@

Merely publishing theory in a journal article does not make it open;
to be open, theory should adhere to established open science standards for specification and archival.
We propose to implement theories as digital objects,
and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data". 
They provide information about the nature and content of a digital object and are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposes, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues [-@lamprechtFAIRPrinciplesResearch2019],
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary (minor) changes,
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR digital object,
and we provide worked examples to encourage their adoption.

@@

It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be archived as FAIR theory. 
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind of information is useful.
As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
For example, following Meehl, we could envision a theory that starts out with establishing, through observation, an ontology of constructs relevant for a given phenomenon.
After initial exploratory research, the theory might be further specified by making assumptions about how these constructs are causally connected.
Over time, more precise *statistical/mathematical models* could be derived by further assuming a specific functional form for relationships (e.g., linear effects) and error families for the distribution of measured variables (e.g., normal distributions).
This allows for the specification of statistical models, which make just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Going even further, a *generative/computational model* could be specified,
which is completely parameterized (e.g., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement instruments (e.g., different questionnaires used to assess the same construct), or information about participant recruitment- and retention strategies.

Theories can include or reference other theories.
For example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
such as adherence to infection prevention protocols (e.g., social distancing),
pandemic-related behavior (e.g., panic buying),
and pandemic-related distress
[@taylor2022psychology].
Such a theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.

@@

Crucially for the present paper, formalization is orthogonal to FAIRification.
FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it pertains to rigorous and transparent archival and communication about theories, with the aim of enhancing their reusability.
FAIR theory does not require formalization, and formal theories are not automatically FAIR.
The FAIR principles
apply to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).
This principle is illustrated in our [vignette on FAIRifying De Groot's empirical cycle](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html):
it is equally possible to FAIRify the theory in its original formulation by archiving a text document with five plain-langues propositions,
or to formalize the theory and represent it as a human- and machine-readable diagram before FAIRifying it.

@@

<span id = "modularvalue">The primary unit of scientific communication has long been the academic paper.
Yet scholars often produce many other valuable resources in the process of writing papers,
including instruments, materials, data, code, and theory.
These resources are often merely described in papers and not made available for reuse.
Modular publishing is the practice of making each of these resources available as independent *digital objects*,
facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].
We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
</span>
<span id="datasharing">At the time of writing, some modular publishing practices are already widely adopted;
data sharing, for example, has become the de-facto standard in psychology in the past decade [@tedersooDataSharingPractices2021].</span>

@@

The field of computer science provides inspiration for well-established processes for iteratively improving digital objects.
Version control systems, like Git, are used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete annotated history of those changes.
It has previously been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
For example, Git can facilitate reproducible research, manage distributed collaboration, and improve preregistrations [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
documenting why changes were made,
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
Note that, while cloud archives associated with Git (e.g., GitHub) facilitate collaborative theory development, 
they are not suitable for archiving the version of record due to a lack of FAIR-compliance.
Thus, theory development may take place on GitHub, but versions of record should be archived on a platform like like Zenodo, with appropriate metadata.

@@

Aside from technical solutions, version control is a social process as well.
On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
<!-- NvD: Maybe we can add an  example (e.g., the theory that explains ego-depletion and the failed replication of the phenomenon). -->
In computer science, these challenges are navigated by assigning version numbers.
Specifically, *semantic versioning* comprises a simple set of rules for assigning version numbers to digital objects.
Whereas version control tracks changes,
semantic versioning communicates what those changes mean to users of the theory,
guides the social process of theory development, and signals how much a theory has been changed.

We propose adaptating semantic versioning for theories by assigning a version number in the format `MAJOR.MINOR.PATCH` (e.g., 0.1.0), where the `MAJOR` number is incremented when backwards incompatible changes are made. For example, if the theory now contains empirical statements that are at odds with a previous version of the theory. The `MINOR` number should be incremented when the set of empirical statements are expanded in a backward compatible manner (i.e., the previous version is subsumed within the new version). The `PATCH` number should be incremented when making backward compatible bug fixes, cosmetic changes, fix spelling errors, or add clarifications.

@@

The four Findability criteria are applicable to theory with only minor adjustments, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for archived objects,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata	(F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
ontology [@guyonMeasurementOntologyEpistemology2018],
or catalog of psychological phenomena.
Metadata should also include identifiers for all the versions of the theory it describes	(F3);
Zenodo handles this by default by providing an overarching DOI for a digital object which subsumes the DOIs of its subversions.

Finally, metadata should be registered or indexed in a searchable registry (F4).
It is important to note that, while many archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
only few are specifically designed for FAIR-compliant archival.
Zenodo stands out in this respect.
Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a specific) FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

Findability is substantially amplified if intended users of a resource know where to search for it.
This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
Regrettably, most academic search engines only index traditional print publications, not other digital objects.
Since the status quo is to publish theories in papers,
the FAIR requirements are met if scholars continue to do so,
and additionally publish theories as separate digital objects.
The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
<!-- An ad-hoc solution is to publish a paper as documentation for the digital object [@mcgillivrayDeepImpactStudy2022]. -->
<!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
<!-- The same solution could be applied to theories. -->
In the longer term, it may not be necessary to write a paper for each theory.
If Zenodo becomes more recognized as a centralized repository for digital objects, researchers may begin to search there more regularly.
Conversely, as organizations (e.g., Google Scholar, Web of Science, Pure, ORCID) begin to recognize other forms of academic output beyond papers, they may begin to index digital objects stored in Zenodo.

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and posted many examples on [their website](https://www.theorymaps.org) [@grayHowMapTheory2017].
Similarly, [PsychoModels](http://psychomodels.org/) seeks to inventorize theories and models in psychology [@vandongenPsychoModelsMathematicalComputational2025].
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
The data engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.
